---
title: "Publication Figure 6"
output:
  html_notebook:
    code_folding: hide
    theme: united
    toc: yes
    toc_depth: 2
    toc_float: true
    toc_collapsed: false
theme: cosmo
---


# Libraries
Standard Import
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(scales)
library(ggpubr)
```

Special
```{r}
# devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
# library(rgr)
# library(mixOmics)
library(pairwiseAdonis)
library(grid)
library(factoextra)
library(FactoMineR)
```

# Paths
```{r}
bin_dir = '../bin/'
data_dir = '../data/'
results_dir = '../results/'
```

# Custom Scripts / Theme
```{r}
source(paste0(bin_dir, 'theme_settings.R'))
source (paste0(bin_dir, 'utilities.R'))
source(paste0(bin_dir, 'species_competition_functions.R'))
source(paste0(bin_dir, 'distmat_utils.R'))
source(paste0(bin_dir, 'analysis_metadata.R'))
date <- format(Sys.time(), "%d%m%y")
```

# Import Tables
Metadata
```{r message=FALSE, warning=FALSE}
samples.metadata <- read_tsv(paste0(data_dir, 'samples.metadata.tsv'))
microbe.taxonomy <- read_tsv(paste0(data_dir, 'microbe.taxonomy.tsv'))
microbe.metadata <- microbe.taxonomy %>% 
  right_join(., read_tsv(paste0(data_dir, 'microbe.metadata.tsv'))) %>% 
  mutate(gram.bool = ifelse(gram_stain == 'positive', T, 
                     ifelse(gram_stain == 'negative', F, NA)), 
         spores.bool = ifelse(spore_forming == 'spore-forming', T, 
                       ifelse(spore_forming == 'non-spore-forming', F, NA))) 

```

## Taxonomy
MetaPhlAn2 tables
combine with taxonomy
```{r message=FALSE, warning=FALSE}
mp_species.long <- microbe.taxonomy %>% 
  right_join(., read_tsv(paste0(data_dir, 'samples.mp_profiles.species.tsv')), 
             by = 'species') %>% 
  left_join(., samples.metadata, by = 'Name')
```

rCDI subset
annotate with metadata
```{r}
mp_species.rcdi.long <- 
  mp_species.long %>%
  filter(Study %in% c(rcdi.studies)) 
```

## Case Summary
SameStr Case-Summary table
```{r message=FALSE, warning=FALSE}
sstr_cases <- read_tsv(paste0(data_dir, 'samples.case_summary.tsv')) %>% 
  left_join(., microbe.taxonomy, by = 'species')
```

# Set Figure
```{r}
fig = paste0('Fig_6.')
```

# Failed/Resolved Symptoms
Focussing on the difference between successful FMT cases and cases in which symptoms were reported to no have been resolved after FMT treatment.

## %Recipient/Donor-derived
### Format Data
```{r fig.width=5, fig.height=5, message=FALSE, warning=FALSE, paged.print=FALSE}
sstr_cases.rcdi.metrics <-
  sstr_cases %>% 
  
  filter(Study %in% c('ALM', 'FRICKE')) %>% 
  filter(Case_Name %in% cases.full) %>% 

  tag_species_competition(.) %>% 

  mutate(n = 1) %>% 
  mutate(source = ifelse(grepl(species, pattern = 'unclassified'),  'Unclassified', source)) %>% 
  
  mutate(source = case_when(
  analysis_level == 'species' & source == 'self' ~ 'Self Sp.',
  analysis_level == 'species' & source == 'donor' ~ 'Donor Sp.',
  analysis_level == 'species' & source == 'unique' ~ 'Unique Sp.',
  T ~ source
  )) %>% 
  
  group_by(Study_Type, Case_Name, source, Days_Since_FMT.post, fmt_success.label) %>% 
  summarize(n = sum(n, na.rm = T),
            rel_abund = sum(rel_abund.post, na.rm = T) / 100) %>% 
  group_by(Case_Name, Days_Since_FMT.post) %>% 
  mutate(f = n / sum(n, na.rm = T)) %>%
  ungroup() %>% 

  group_by(Case_Name) %>% 
  mutate(source = case_when(source == 'Unclassified' ~ 'Unclassified Sp.',
                  source == 'same_species' ~ 'Same Sp.', 
                  source == 'unique' ~ 'Unique to Time Point',
                  source == 'self' ~ 'Recipient / Initial Sample',
                  source == 'donor' ~ 'Donor',
                  source == 'both' ~ 'Coexistence', 
                  T ~ source)) %>% 
  
  pivot_wider(names_from = 'source', 
               values_from = c('rel_abund','f', 'n'), names_sep = '___') %>% 
  mutate_at(.vars = vars(contains('___')),
            .funs = funs(replace_na(., 0))) %>%
  pivot_longer(cols = contains("___"), 
               names_to = c("metric", "source"),
               names_sep = '___', values_drop_na = F) %>% 
  mutate(fmt_success.label = ifelse(fmt_success.label == 'Failed', 
                                    'Unresolved Symptoms', 'Resolved Symptoms'))
```

```{r}
sstr_cases.rcdi.rel_abund.annotated <-
  sstr_cases.rcdi.metrics %>% 
  filter(metric == 'rel_abund') %>% 
  filter(Days_Since_FMT.post <= 37) %>% 
  
  # keep only recipient and donor-derived (excluding coexistence)
  mutate(
    source.simple = case_when(
           source %in% c('Recipient / Initial Sample','Self Sp.') ~ 'Recipient',
           source %in% c('Donor','Donor Sp.') ~ 'Donor',
           T ~ 'Drop'
         )) %>% 
  filter(source.simple != 'Drop') %>% 
  
  group_by(Case_Name, Days_Since_FMT.post, source.simple, fmt_success.label) %>% 
  summarize(value = sum(value, na.rm = T)) %>% 
  ungroup() 
```


### Scatterplot over time
Fraction of donor and recipient-derived species over time for failed and resolved cases
```{r fig.width=6, fig.height=3}
plot <-
  sstr_cases.rcdi.rel_abund.annotated %>% 

  
  ggplot(aes(Days_Since_FMT.post, value)) +
  geom_point(aes(fill = source.simple), alpha = 0.5, 
             shape = 21, size = 3, col = 'black') + 
  geom_line(aes(col = source.simple, 
                group = paste0(Case_Name, source.simple)
                ), show.legend = F, alpha = 0.5) +
  geom_smooth(se = F, method = 'glm', aes(col = source.simple),
              method.args=list(family='binomial')) +
  geom_hline(yintercept = 0.25, linetype = 'dashed', size = .5) + 

  scale_color_manual(values = colors.discrete[c(1,3)]) + 
  scale_fill_manual(values = colors.discrete[c(1,3)]) + 
  scale_y_continuous(labels = percent_format_signif, 
                     breaks = c(0, .25, .5, .75, 1), 
                     limits = c(0, 1), expand = c(0,0)) +
  scale_x_continuous(trans = pseudo_log_trans(0.5, 10), 
                     breaks = c(2, 7, 14, 35, 84, 365)) + 
  facet_grid(cols = vars(fmt_success.label)) + 
  theme_cowplot() +
  theme(axis.title.x = element_blank(),
        axis.line = element_line(size = 0.2, color = 'black'), 
        strip.text.x = element_text(size = 14),
        strip.background = element_blank()) + 
  panel_border(remove = F, size = 1, color = 'black') +
  labs(y = 'Post-FMT Sample\nRel. Abund. by Source') +
  guides(fill = guide_legend(title = 'Post-FMT\nMicrobiota Source', ncol = 1))

plot + theme(legend.position = 'none')
legend = cowplot::get_legend(plot)
grid.newpage()
grid.draw(legend)
```

Exporting plot
```{r}
output_name = 'PostSource.Cases.Time.rCDI'

ggsave(plot + theme(legend.position = 'none'), 
       device = 'pdf', dpi = 300, width = 6, height = 3,
       filename = paste0(results_dir, fig, output_name, '.Plot.pdf'))
ggsave(legend, 
       device = 'pdf', dpi = 300, width = 3, height = 3,
       filename = paste0(results_dir, fig, output_name, '.Legend.pdf'))
```


Persistence / Engraftment within time periods:
```{r}
sstr_cases.rcdi.rel_abund.annotated %>% 
  filter(Days_Since_FMT.post > 0, Days_Since_FMT.post <= 7) %>% 
  group_by(source.simple, fmt_success.label) %>% 
  summarize(mean = mean(value) * 100, 
            sd = sd(value) * 100, 
            .groups = 'drop')

sstr_cases.rcdi.rel_abund.annotated %>% 
  filter(Days_Since_FMT.post > 7, Days_Since_FMT.post <= 37) %>% 
  group_by(source.simple, fmt_success.label) %>% 
  summarize(mean = mean(value) * 100, 
            sd = sd(value) * 100, 
            .groups = 'drop')
```
Resolved:
Donor -> 44.8 ± 30.1  to  52.0 ± 23.6
Recipient -> 20.9 ± 24.9  to  12.5 ± 17.0

Failed:
Donor -> 33.1 ± 23.5  to  24.3 ± 8.7
Recipient -> 25.2 ± 19.0  to  40.1 ± 24.3

### Boxplot
Boxplots at <=1W, 1-6W, >6W
```{r}
sstr_cases.rcdi.metrics %>% 
  filter(metric == 'rel_abund') %>% 
  mutate(Days_Since_fmt_success.label = case_when(
    Days_Since_FMT.post <7 ~ '<1W', 
    Days_Since_FMT.post >=7 & Days_Since_FMT.post <= 42 ~ '1-6W',
    Days_Since_FMT.post > 42 ~ '>6W')) %>% 

  # keep only recipient and donor-derived (excluding coexistence)
  mutate(
    source.simple = case_when(
           source %in% c('Recipient / Initial Sample','Self Sp.') ~ 'Recipient',
           source %in% c('Donor','Donor Sp.') ~ 'Donor',
           
           T ~ 'Drop'
         )) %>% 
  filter(source.simple != 'Drop') %>% 
  
  group_by(Case_Name, Days_Since_FMT.post, Days_Since_fmt_success.label, source.simple, fmt_success.label) %>% 
  summarize(value = sum(value, na.rm = T)) %>% 
  ungroup() %>% 
  
  mutate(Days_Since_fmt_success.label = fct_relevel(Days_Since_fmt_success.label, '<1W', '1-6W', '>6W')) %>% 
  
  ggplot(aes(Days_Since_fmt_success.label, value, fill = source.simple)) + 
  geom_boxplot() +
  scale_y_continuous(labels = percent_format_signif) +
  facet_grid(cols = vars(gsub(fmt_success.label, pattern = ' ', replacement = '\n')), 
             space = 'free', scales = 'free_x') + 
  theme_cowplot() + 
  theme(axis.title.x = element_blank(),
        strip.background = element_blank()) + 
  scale_fill_manual(values = colors.discrete[c(1,3)]) + 
  labs(y = 'Cumulative\nSample Rel. Abund.') + 
  guides(fill = guide_legend(title = 'Post-FMT\nMicrobiota Source', ncol = 2))
```

## Diversity Differences
### Alpha: Shannon
#### Format
```{r}
# transpose
mp_species.wide <- mp_species.long %>%
  filter(kingdom == 'Bacteria',
         Study_Type == 'rCDI') %>% 
  pivot_wider(id_cols = 'species', 
              names_from = 'Name',
              values_from = 'rel_abund', 
              values_fill = list(rel_abund = 0))

# alpha diversity
mp_species.samples <- t(select(mp_species.wide, matches('.pair'), species) %>% column_to_rownames(var = 'species'))
mp_species.shannon <- data.frame(Shannon = diversity(mp_species.samples, index = 'shannon')) %>% rownames_to_column('Name') %>% 
left_join(., samples.metadata, by = 'Name')
```

#### Boxplots
Shannon Index of rCDI and Control data showing that cases in which FMT failed to resolve clinical symptoms can not be distinguished by controls, successfully treated patients or donors in terms of Shannon Index. Index is significantly lower in pre-FMT samples. 
Since MGH03D-Donor (only donor with failed cases) on higher end of Shannon Index, alpha diversity probably not a very relevant indicator for choosing donors.
```{r fig.width=2, fig.height=2.5}
  
ww = 0.5
plot <-
  mp_species.shannon %>% 
  filter(!is.na(Sample_Type)) %>% 
  # filter(last_sample) %>% 
  
  mutate(Sample_Type.label = str_to_upper(unlist(str_extract(Sample_Type, '.')))) %>% 
  mutate(tag = case_when(
    Donor.Subject != 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' ~ Sample_Type.label,
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & 
      (Sample_Type == 'donor' | (Sample_Type == 'post' & !fmt_success)) ~ paste0(Sample_Type.label, '*'),
    T ~ Sample_Type.label
  )) %>% 
  mutate(tag = fct_relevel(tag, 'R', 'P', 'D', 'P*', 'D*')) %>% 
  # filter(tag != 'R') %>% 
  
  ggplot(aes(tag, Shannon)) + 
  stat_boxplot(geom = 'errorbar', width = ww) +
  geom_boxplot(show.legend = F, width = ww) + 
  geom_dotplot(aes(fill = NULL), binaxis = "y", 
               stackdir = "center", binpositions="all", binwidth = 0.05) +
  
  theme_cowplot() + 
  theme(axis.title.x = element_blank()) + 
  labs(y = 'Shannon Index') + 
  annotate('text', label = '*Unresolved\nSymptoms', y = 0.45, x = 'D*', hjust = 1) # + 
  # stat_compare_means(method = 'kruskal')

plot
```

Exporting plot
```{r}
output_name = 'TaxonomicDiversity.Shannon'

ggsave(plot + theme(legend.position = 'none'), 
       device = 'pdf', dpi = 300, width = 2, height = 2.5,
       filename = paste0(results_dir, fig, output_name, '.Plot.pdf'))
```

### Beta: PCA of species-level taxonomic composition
PCA of clr-transformed taxonomic composition

#### Diversity Calculations
```{r}
# transpose
mp_species.rcdi.wide <- mp_species.rcdi.long %>%
  filter(kingdom == 'Bacteria') %>%
  pivot_wider(id_cols = 'species', 
              names_from = 'Name',
              values_from = 'rel_abund', 
              values_fill = list(rel_abund = 0))

## -- genus-level
mp_genus.rcdi.wide <- mp_species.rcdi.long %>%
  filter(kingdom == 'Bacteria') %>% 
  pivot_wider(id_cols = 'genus', 
              names_from = 'Name',
              values_from = 'rel_abund', 
              values_fill = list(rel_abund = 0), 
              values_fn = list(rel_abund = sum))

# get data
mp_species.rcdi.data <- t(select(mp_species.rcdi.wide, 
                                    matches('.pair'), species) %>% 
                               column_to_rownames(var = 'species'))

## -- genus
mp_genus.rcdi.data <- t(select(mp_genus.rcdi.wide, 
                                    matches('.pair'), genus) %>% 
                               column_to_rownames(var = 'genus'))

# species-level
pseudo = 1e-6
mp_species.rcdi.clr <- rgr::clr(mp_species.rcdi.data + pseudo,  ifclose = FALSE, ifwarn = TRUE)

# aitchison dist
mp_species.rcdi.clr.dist <- dist(scale(mp_species.rcdi.clr, center = T, scale = F), 
                                 method = 'euclidean') %>% as.matrix()

# pca
mp_species.rcdi.clr.pca_data <- PCA(mp_species.rcdi.clr, graph = FALSE, scale.unit = F)
mp_species.rcdi.clr.pca <- mp_species.rcdi.clr.pca_data %>% 
  .$ind %>% .$coord %>% 
  as.data.frame() %>% 
  rownames_to_column(var = 'Name') 

## annotate with metadata, diversity
mp_species.rcdi.data.clr.pca <- mp_species.rcdi.clr.pca %>% 
  left_join(., samples.metadata, by = c('Name')) %>% 
  left_join(., mp_species.shannon %>% select(Name, Shannon), by = c('Name')) %>% 
  group_by(Case_Name, Sample_Type) %>% 
  mutate(last = dense_rank(Days_Since_FMT) == max(dense_rank(Days_Since_FMT))) %>% 
  mutate(d3.tag = case_when(
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & Sample_Type == 'post' ~ 
      paste0('post | D3 ', if_else(fmt_success, 'Success', 'Failed')),
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & Sample_Type == 'donor' ~ 'donor | D3',
    T ~ Sample_Type)) %>% 
  ungroup()
```

#### D/P Difference
PERMANOVA (adonis function) testing between rCDI Donors & Post-FMT: MGH03D with failed cases (3 samples) and other donors.
```{r}
# donor & post-FMT
selection <- mp_species.rcdi.long %>% 
  filter(Sample_Type %in% c('post', 'donor'), last_sample) %>% 
  distinct(Name, Donor.Subject, Sample_Type, fmt_success)

d3.adonis <-
  pairwiseAdonis::pairwise.adonis(
  x = mp_species.rcdi.clr[selection$Name, ],
  factors = ifelse(selection$Donor.Subject == "D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21", 'MGH03D-related','Other'),
  sim.method = 'euclidean',
  perm = 1e3) %>% 
  rename(fdr = p.adjusted) %>% 
  mutate(sig = cut(fdr, 
                   breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                   labels = c("***", "**", "*", "ns")))

d3.adonis
```

#### PCA biplot

##### Format
Abbr. species names in PCA object
```{r}
select.prevotella <- c('Prevotella_denticola', 'Prevotella_oris', 'Prevotella_oralis','Prevotella_veroralis')
select.fusobacteria <- c('Fusobacterium_nucleatum', 'Fusobacterium_periodonticum', 'Fusobacterium_mortiferum')
other <-  c('Rothia_dentocariosa', 'Veilonella_parvula')

select.bacteroides <- c('Bacteroides_faecis', 'Bacteroides_eggerthii', 'Bacteroides_intestinalis', 'Bacteroides_thetaiotaomicron') 
# 'Bacteroides_fragilis',
select.ruminococcus <- c('Ruminococcus_torques', 'Ruminococcus_gnavus')
# 'Ruminococcus_albus', 'Ruminococcus_callidus', 'Bacteroides_obeum'

plot_spp <- c(other, select.bacteroides, select.ruminococcus, select.prevotella, select.fusobacteria,
               'Lachnospiraceae_bacterium')

print_taxonomy <- function(name, selection) {
  unlist(
    lapply(name, function(x) {
      case_when(
        !x %in% selection ~ x,
        x %in% selection ~ str_replace(str_replace(x, '_', ' '), '[a-z].*\ ', '. ')
      )
    }
    )) %>% 
    return()
}

rownames(mp_species.rcdi.clr.pca_data$var$coord) <- print_taxonomy(rownames(mp_species.rcdi.clr.pca_data$var$coord), plot_spp)
rownames(mp_species.rcdi.clr.pca_data$var$cor) <- print_taxonomy(rownames(mp_species.rcdi.clr.pca_data$var$cor), plot_spp)
rownames(mp_species.rcdi.clr.pca_data$var$cos) <- print_taxonomy(rownames(mp_species.rcdi.clr.pca_data$var$cos), plot_spp)
rownames(mp_species.rcdi.clr.pca_data$var$contrib) <- print_taxonomy(rownames(mp_species.rcdi.clr.pca_data$var$contrib), plot_spp)
```

```{r}
# https://stackoverflow.com/questions/13407236/remove-a-layer-from-a-ggplot2-chart
remove_geom <- function(ggplot2_object, geom_type) {
  # Delete layers that match the requested type.
  layers <- lapply(ggplot2_object$layers, function(x) {
    if (class(x$geom)[1] == geom_type) {
      NULL
    } else {
      x
    }
  })
  # Delete the unwanted layers.
  layers <- layers[!sapply(layers, is.null)]
  ggplot2_object$layers <- layers
  ggplot2_object
}
```

```{r fig.height=5,fig.weigth=5}
pp <-
  factoextra::fviz_pca_biplot(mp_species.rcdi.clr.pca_data, 
             # samples
             fill.ind = mp_species.rcdi.data.clr.pca$d3.tag, 
             pointshape = 21, 
             pointsize = 8, 
             alpha.ind = ifelse(grepl(mp_species.rcdi.data.clr.pca$d3.tag, pattern = 'D3'), 1, .25),
             
             # features
             select.var = list('name' = print_taxonomy(plot_spp, plot_spp)),
             label = 'var', labelsize = 3, 
             col.var = 'black',  
             repel = T, invisible = c("quali"), 
             
             # general
             addEllipses = F, centroids = F, scale = F,
             title = '',
             subtitle = 'Taxonomic Composition',
             )  + 

  ggpubr::fill_palette(palette = c(colors.discrete[c(1,1,2)], 
                                   '#B4907E',  #A47963
                                   colors.discrete[c(2,3)])) +
  scale_x_reverse() + 
  scale_y_reverse() +
  theme_cowplot() + 
  theme(aspect.ratio = 1,
        plot.subtitle = element_text(hjust = .5, size = 12), 
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 8),
)  + 
  coord_flip()

# axes - rename
pp$labels$x <- str_replace(pp$labels$x, pattern = 'Dim', replacement = 'PC ')
pp$labels$y <- str_replace(pp$labels$y, pattern = 'Dim', replacement = 'PC ')

# arrows - rescale
pp[["layers"]][[5]][["data"]][["x"]] = pp[["layers"]][[5]][["data"]][["x"]]  *.4
pp[["layers"]][[5]][["data"]][["y"]] = pp[["layers"]][[5]][["data"]][["y"]]  *.4

# labels
pp[["layers"]][[4]][["data"]][["x"]] = pp[["layers"]][[4]][["data"]][["x"]]  *.4
pp[["layers"]][[4]][["data"]][["y"]] = pp[["layers"]][[4]][["data"]][["y"]]  *.4

# crosshair
pp <- remove_geom(pp, 'GeomHline')
pp <- remove_geom(pp, 'GeomVline')

pp <- pp +
  annotate('text', x = -52, y = 52, hjust = 0, vjust = 1, size = 3, 
           label = paste0('PERMANOVA (Post-FMT & Donors)\n', 
                         'MGH03D-related vs. Others\n')) + 
  annotate('text', x = -41.5, y = 52, hjust = 0, vjust = 1, size = 3,
           label = bquote(R^2 * '=' *  .(format(100*d3.adonis$R2,digits=3))*'%' * .(format(d3.adonis$sig))))

legend <- cowplot::get_legend(pp)

pp + theme(legend.position = 'none')
grid.newpage()
grid.draw(legend)
```


Exporting plot
```{r}
output_name = 'TaxonomicDiversity.rCDI.CLR_PCAbiplot'

ggsave(pp + theme(legend.position = 'none'), 
       device = 'pdf', dpi = 300, width = 5, height = 5,
       filename = paste0(results_dir, fig, output_name, '.Plot.pdf'))
ggsave(legend, 
       device = 'pdf', dpi = 300, width = 5, height = 5,
       filename = paste0(results_dir, fig, output_name, '.Legend.pdf'))
```

#### Donor Difference
```{r}
# donor only
donors <- mp_species.rcdi.long %>% 
  filter(Sample_Type %in% c('donor')) %>% 
  distinct(Name, Donor.Subject)

selection <- donors %>% 
  pull(Name)

mp_species.rcdi.clr.pca_data <- PCA(mp_species.rcdi.clr[selection, ], graph = FALSE, scale.unit = F)

# get data
mp_species.rcdi.data <- t(select(mp_species.rcdi.wide, 
                                    matches('.pair'), species) %>% 
                               column_to_rownames(var = 'species'))

# species-level
pseudo = 1e-6
mp_species.rcdi.donors.data <- mp_species.rcdi.data[selection, colSums(mp_species.rcdi.data[selection,]) > 0]
mp_species.rcdi.donors.clr <- rgr::clr(mp_species.rcdi.donors.data + pseudo,  ifclose = FALSE, ifwarn = TRUE)

# PCA
mp_species.rcdi.donors.clr.pca_data <- PCA(mp_species.rcdi.donors.clr, graph = FALSE, scale.unit = F, ncp = 2)
mp_species.rcdi.donors.clr.pca <- mp_species.rcdi.donors.clr.pca_data %>% 
  .$ind %>% .$coord %>% 
  as.data.frame()
mp_species.rcdi.donors.clr.expl <- mp_species.rcdi.donors.clr.pca_data$eig[1:2,'percentage of variance'] / 100


## annotate with metadata, diversity
mp_species.rcdi.donors.data.clr.pca <- mp_species.rcdi.donors.clr.pca %>% 
  rownames_to_column('Name') %>% 
  left_join(., samples.metadata, by = c('Name')) %>% 
  left_join(., mp_species.shannon %>% select(Name, Shannon), by = c('Name')) %>% 
  group_by(Case_Name, Sample_Type) %>% 
  mutate(last = dense_rank(Days_Since_FMT) == max(dense_rank(Days_Since_FMT))) %>% 
  mutate(d3.tag = case_when(
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & Sample_Type == 'post' ~ 
      paste0('post | D3 ', if_else(fmt_success, 'Success', 'Failed')),
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & Sample_Type == 'donor' ~ 'donor | D3',
    T ~ Sample_Type)) %>% 
  ungroup()
```

Pairwise PERMANOVA (adonis function) testing between rCDI Donors: MGH03D with failed cases (3 samples) and other donors.
```{r}
d3.adonis <-
  pairwiseAdonis::pairwise.adonis(
  x = mp_species.rcdi.clr[donors$Name, ],
  factors = ifelse(donors$Donor.Subject == "D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21", 'MGH03D','Other'),
  sim.method = 'euclidean',
  p.adjust.m = 'fdr',
  perm = 1e3) %>% 
  mutate(R = sqrt(R2)) %>% 
  rename(p.bonferroni = p.adjusted)
d3.adonis
```

```{r fig.width=2, fig.height=3}
# aitchison dist
mp_species.rcdi.donors.clr.dist <- dist(scale(mp_species.rcdi.donors.clr, center = T, scale = F), 
                                 method = 'euclidean') %>% as.matrix()


mp_species.rcdi.donors.clr.dist.long <- mp_species.rcdi.donors.clr.dist %>% 
  as.data.frame(.) %>% 
  distmat_to_long(., value_name = 'dist', rm_diag = T) %>%
  left_join(., samples.metadata %>% select(Name, Donor.Unique_ID, Donor.Subject), 
            by = c('row' = 'Name')) %>% 
  left_join(., samples.metadata %>% select(Name, Donor.Unique_ID, Donor.Subject), 
            by = c('col' = 'Name'), suffix = c('.row','.col')) %>% 
  mutate(d3.tag = Donor.Subject.row == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' | 
                  Donor.Subject.col == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' 
                  ) %>% 
  filter(Donor.Subject.row != Donor.Subject.col)
```


```{r fig.width=5, fig.height=3}
mp_species.rcdi.donors.clr.dist.long %>% 
  mutate(d3.tag = ifelse(d3.tag, 'MGH03D\nOther','Other\nOther')) %>% 
  ggplot(aes(x = (dist - mean(mp_species.rcdi.donors.clr.dist.long$dist)) / 
                    sd(mp_species.rcdi.donors.clr.dist.long$dist), 
             fill = d3.tag)) + 
  geom_density(alpha = 0.5) + 
  scale_y_continuous(breaks = c(0, .25, .5),
                     labels = percent_format()) +
  geom_vline(xintercept = 1, linetype = 'dashed', alpha = 0.5) +
  theme_cowplot() + 
  theme(aspect.ratio = 1,
        legend.position = 'top',
        legend.title = element_blank()) + 
  labs(x = 'Beta-Diversity (Z-Score)') 
```


```{r}
output_name = 'TaxonomicDiversity.CLR_Euc.MGH03D.'
ggsave(plot + theme(legend.position = 'none'),
       device = 'pdf', dpi = 300, width = 2.5, height = 4,
       filename = paste0(results_dir, fig, output_name, '.Plot.pdf'))
```

# Taxa Distinguishing MGH03D-related / Other

## Format
```{r}
mp_species.rcdi.annotated <- 
  samples.metadata %>% 
  mutate(tag = paste0(Unique_ID, ' ', gsub(
      gsub(Name, pattern = '.*_', replacement = ''), 
                 pattern = '.pair', replacement = ''))) %>% 
  right_join(., as.data.frame(mp_species.rcdi.data) %>% rownames_to_column('Name')) %>% 
  filter(Sample_Type %in% c('post','donor'))

X.metadata <-
  mp_species.rcdi.annotated %>% 
  select(Study:tag) %>% 
  mutate(Y = ifelse(Donor.Subject == "D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21", 'MGH03D-related', 'Other'))

X = 
  mp_species.rcdi.annotated %>% 
  column_to_rownames(var = 'tag') %>% 
  select(-Study:-Study_Type) %>% 
  as.data.frame()

Y = as.factor(X.metadata$Y)
```

## sparse PLS Discriminant Analysis

### Initial sPLS-DA
sPLS-DA of rCDI recipients and donor samples based on taxonomy
```{r}
pseudo = 1e-6
data.plsda = mixOmics::plsda(X = X + pseudo, Y, ncomp = nlevels(Y), logratio = 'CLR')
data.perf.plsda = mixOmics::perf(data.plsda, validation = 'Mfold', folds = 5,
                    progressBar = FALSE, nrepeat = 10)

plot(data.perf.plsda, overlay = 'measure', sd=TRUE)
mixOmics::plotIndiv(data.plsda , comp = c(1,2),
          group = Y, 
          ind.names = T,
          ellipse = T,
          legend = TRUE, title = 'PLSDA comp 1 - 2')
```

### Cross-Validate features
5x10 CV
```{r}
seed = 100
set.seed(seed)
data.tune.splsda = mixOmics::tune.splsda(X + pseudo, 
                          Y = Y, 
                          ncomp = 2, 
                          multilevel = NULL, 
                          logratio = 'CLR',
                          validation = c('Mfold'), 
                          folds = 5, 
                          dist = 'max.dist', 
                          nrepeat = 10,
                          progressBar = FALSE)

plot(data.tune.splsda)
select.keepX = data.tune.splsda$choice.keepX[1:2]
select.keepX
```

### Apply CV features
```{r}
data.splsda = mixOmics::splsda(X = X + pseudo,  Y = Y, 
                               ncomp = 2, 
                               keepX = select.keepX, 
                               logratio= "CLR",
                               near.zero.var = T,
)
data.perf.splsda = mixOmics::perf(data.splsda, validation = 'Mfold', folds = 5, 
                   progressBar = FALSE, nrepeat = 10, dist = 'max.dist')

data.perf.splsda$error.rate

mixOmics::plotIndiv(data.splsda , comp = c(1,2),
          group = Y, ind.names = T, 
          ellipse = TRUE, legend = TRUE, title = 'PLSDA comp 1 - 2')
```

Plot loadings and save to table
```{r}
# COMP 1
pL.pc1 <- mixOmics::plotLoadings(data.splsda, title = 'sPLS-DA PC 1',
                       comp = 1, method = 'median', contrib = 'max', 
                       size.title = rel(1), border = T, 
                       size.name = .5, size.legend = .75, 
                       legend.color = colors.discrete[c(4, 1)])
pL.pc1

# COMP 2
pL.pc2 <- mixOmics::plotLoadings(data.splsda, title = 'sPLS-DA PC 2',
                       comp = 2, method = 'median', contrib = 'max', 
                       size.title = rel(1), border = T, 
                       size.name = .5, size.legend = .75, 
                       legend.color = colors.discrete[c(4, 1)])
pL.pc2

pL <- 
  bind_rows(pL.pc1, pL.pc2) %>% 
  rownames_to_column('species') %>% 
  left_join(., microbe.metadata, by = 'species') %>% 
  rename(rel_abund.highest = GroupContrib) %>% 
  select(species, 
         habit.site,
         oxygen.class,
         MGH03D.related, 
         Other, 
         rel_abund.highest, 
         importance)

output_name = 'sPLSDA.MGH03D.Taxonomic.Loadings'
write_tsv(pL, paste0(results_dir, fig, output_name, '.tsv'))
```

Heatmap plot of differentially abundant taxa 
MGH03D-related:
+ Prevotella and Bacteroides/Parabacteroides-related species 
+ Odoribacter, Megasphaera, Alistipes, Desulfovibrio, Bacteroidales, Fusobacterium (mortiferum), Megamonas
- R. torques, Lachnospiraceae, some Eubacterium species
- Lachnospiraceae, Eggerthella, Anaerostipes, Streptococci

This is supported by strain-level transfer showing that in MGH03D-related cases Prevotella, Parabacteroides, Megasphaera, Alistipes and Odoribacter are transferred
Link to sulfate metabolism? IBD?
```{r}
sample_colors <-
  tibble(tag = data.splsda$names$sample) %>% 
  left_join(., X.metadata) %>% 
  mutate(color = 
        case_when(Y == 'MGH03D-related' & Sample_Type == 'donor' ~ colors.discrete[1],
                  Y != 'MGH03D-related' & Sample_Type == 'donor' ~ colors.discrete[6],
                  
                  Y == 'MGH03D-related' & Sample_Type == 'post' & fmt_success ~ colors.discrete[2],
                  Y == 'MGH03D-related' & Sample_Type == 'post' & !fmt_success ~ '#B4907E',
                  
                  Y != 'MGH03D-related' ~ colors.discrete[7], 
                  T ~ 'white'))


output_name = 'sPLSDA.MGH03D.Taxonomic.Heatmap'
mixOmics::cim(data.splsda, row.sideColors = sample_colors$color, 
              symkey = T, keysize = c(0.8, 0.8), 
              row.names = F, 
              col.names = gsub(data.splsda$names$colnames$X, pattern = '_noname', replacement = ''),
              save = 'pdf', 
              name.save = paste0(results_dir, fig, output_name),
              margins = c(25, 10),
              row.cex = .5, 
              transpose = T,
              scale = T, center = T)
```



# Functions Distinguishing MGH03D-related / Other

## Format
```{r}
h2_unstratified <- samples.metadata %>% 
  right_join(., read_tsv(paste0(data_dir, 'samples.h2_profiles.unstratified.tsv')), 
             by = 'Name')
```

Annotate Humann2 with PWY Classes
```{r}
h2.rcdi.wide <-
  h2_unstratified %>% 
  filter(!is.na(pathway_description)) %>%
  reshape2::dcast(., pathway_id + pathway_description ~ Name,
                  value.var = 'CPM', fun.aggregate = sum, na.rm = T)

# transpose
h2.rcdi.wide <- h2_unstratified %>%
  filter(!is.na(pathway_description)) %>%
  pivot_wider(id_cols = 'pathway_description', 
              names_from = 'Name',
              values_from = 'CPM', 
              values_fill = list(CPM = 0))

# get data
h2.rcdi.data <- t(select(h2.rcdi.wide, 
                                    matches('.pair'), pathway_description) %>% 
                               column_to_rownames(var = 'pathway_description'))
```


## Diversity
Diversity metrics based on species-level relative abundances

### Alpha: Shannon Index
Alpha Diversity
```{r}
h2.rcdi.shannon <- data.frame(Shannon = diversity(h2.rcdi.data, index = 'shannon')) %>% rownames_to_column('Name')
```

### CLR-Transform
```{r}
pseudo = 1e-3 # CPM
h2.rcdi.clr <- rgr::clr(h2.rcdi.data + pseudo,  ifclose = FALSE, ifwarn = TRUE)
```

### Beta: PCA of functional composition
```{r}
h2.rcdi.clr.pca_data <- PCA(h2.rcdi.clr, graph = FALSE, scale.unit = F)
h2.rcdi.clr.pca <- h2.rcdi.clr.pca_data %>% 
  .$ind %>% .$coord %>% 
  as.data.frame() %>% 
  rownames_to_column(var = 'Name') 

## annotate with metadata, diversity
h2.rcdi.data.clr.pca <- h2.rcdi.clr.pca %>% 
  left_join(., samples.metadata, by = 'Name') %>% 
  left_join(., h2.rcdi.shannon, by = 'Name') %>% 
  group_by(Case_Name, Sample_Type) %>% 
  mutate(last = dense_rank(Days_Since_FMT) == max(dense_rank(Days_Since_FMT))) %>% 
  mutate(d3.tag = case_when(
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & Sample_Type == 'post' ~ 
      paste0('post | D3 ', if_else(fmt_success, 'Success', 'Failed')),
    Donor.Subject == 'D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21' & Sample_Type == 'donor' ~ 'donor | D3',
    T ~ Sample_Type)) %>% 
  ungroup()
```

```{r}
plot <-
  h2.rcdi.data.clr.pca %>% 

  ggplot(aes(Dim.2, Dim.1, label = Unique_ID)) +
  geom_point(shape = 21, col = 'black', size = 8, 
  aes(fill = Sample_Type)) + # Sample_Type 
  scale_fill_manual(values = c(colors.discrete[c(1,2,3)]), 
                    guide = guide_legend(reverse = TRUE)) +
  theme_cowplot() + 
  labs(subtitle = 'Functional Composition') + 
  coord_fixed() +
  theme(plot.subtitle = element_text(hjust = 0.5),
        legend.title=element_blank(),
        aspect.ratio = 1
        ) + 
  labs(y = paste0('PC 1 (', round(h2.rcdi.clr.pca_data$eig[1,'percentage of variance'], 1), '%)'), 
       x = paste0('PC 2 (', round(h2.rcdi.clr.pca_data$eig[2,'percentage of variance'], 1), '%)'))

legend <- cowplot::get_legend(plot)
plot + theme(legend.position = 'none')
grid.newpage()
grid.draw(legend)
```


## Format
```{r}
h2.rcdi.annotated <- 
  samples.metadata %>% 
  mutate(tag = paste0(Unique_ID, ' ', gsub(
      gsub(Name, pattern = '.*_', replacement = ''), 
                 pattern = '.pair', replacement = ''))) %>% 
  right_join(., as.data.frame(h2.rcdi.data) %>% rownames_to_column('Name')) %>% 
  filter(Sample_Type %in% c('post','donor')) %>% 
  mutate(Y = ifelse(Donor.Subject == "D0_ALM_Case_4;5;6;8;9;12;13;15;17;18;20;21", 'MGH03D-related', 'Other'))


X.metadata <-
  h2.rcdi.annotated %>% 
  select(Study:tag, Y)

X = 
  h2.rcdi.annotated %>% 
  column_to_rownames(var = 'tag') %>% 
  select(-Study:-Study_Type, -Y) %>% 
  as.data.frame()

Y = as.factor(X.metadata$Y)
```

## sparse PLS Discriminant Analysis

### Initial sPLS-DA
sPLS-DA of rCDI recipients and donor samples based on taxonomy
```{r}
pseudo = 1e-3
data.plsda = mixOmics::plsda(X = X + pseudo, Y, ncomp = nlevels(Y), logratio = 'CLR')
data.perf.plsda = mixOmics::perf(data.plsda, validation = 'Mfold', folds = 5,
                    progressBar = FALSE, nrepeat = 10)

plot(data.perf.plsda, overlay = 'measure', sd=TRUE)
mixOmics::plotIndiv(data.plsda , comp = c(1,2),
          group = Y, 
          ind.names = T,
          ellipse = T,
          legend = TRUE, title = 'PLSDA comp 1 - 2')
```

### Cross-Validate features
5x10 CV
```{r}
set.seed(seed)
data.tune.splsda = mixOmics::tune.splsda(X + pseudo, 
                          Y = Y, 
                          ncomp = 2, 
                          multilevel = NULL, 
                          logratio = 'CLR',
                          validation = c('Mfold'), 
                          folds = 5, 
                          dist = 'max.dist', 
                          nrepeat = 10,
                          progressBar = FALSE)

plot(data.tune.splsda)
select.keepX = data.tune.splsda$choice.keepX[1:2]
select.keepX
# select.keepX = c(20, 20)
```

### Apply CV features
```{r}
data.splsda = mixOmics::splsda(X = X + pseudo,  Y = Y, 
                               ncomp = 2, 
                               keepX = select.keepX, 
                               logratio= "CLR",
                               near.zero.var = T,
)
data.perf.splsda = mixOmics::perf(data.splsda, validation = 'Mfold', folds = 5, 
                   progressBar = FALSE, nrepeat = 10, dist = 'max.dist')

data.perf.splsda$error.rate

mixOmics::plotIndiv(data.splsda , comp = c(1,2),
          group = Y, ind.names = T, 
          ellipse = TRUE, legend = TRUE, title = 'PLSDA comp 1 - 2')
```

Get colors for sample columns and taxa rows
```{r}
sample_colors <-
  tibble(Unique_ID.splsda = str_split_fixed(data.splsda$names$sample, pattern = ' ', n = 2)[,1]) %>% 

  cbind(., h2.rcdi.annotated) %>%

  mutate(color = 
        case_when(Y == 'Other' & Sample_Type == 'post' ~ colors.discrete[7],
                  Y == 'Other' & Sample_Type == 'donor' ~ colors.discrete[6],
                  
                  Y == 'MGH03D-related' & Sample_Type == 'post' & fmt_success ~ colors.discrete[2],
                  Y == 'MGH03D-related' & Sample_Type == 'post' & !fmt_success ~ '#B4907E',
                  Y == 'MGH03D-related' & Sample_Type == 'donor' ~ colors.discrete[1], 
                  T ~ 'white'))
```


Plot loadings and save to table
```{r fig.width=10,fig.height=10}
# COMP 1
pL.pc1 <- mixOmics::plotLoadings(data.splsda, title = 'sPLS-DA PC 1',
                       comp = 1, method = 'median', contrib = 'max', 
                       size.title = rel(1), border = T, 
                       size.name = .5, size.legend = .75, 
                       legend.color = colors.discrete[c(4, 1)])
pL.pc1

# COMP 2
pL.pc2 <- mixOmics::plotLoadings(data.splsda, title = 'sPLS-DA PC 2',
                       comp = 2, method = 'median', contrib = 'max', 
                       size.title = rel(1), border = T, margins = c(0, 0),
                       size.name = .5, size.legend = .75, 
                       legend.color = colors.discrete[c(4, 1)])
pL.pc2

pL <- 
  bind_rows(pL.pc1, pL.pc2) %>% 
  rownames_to_column('pathways') %>% 
  rename(rel_abund.highest = GroupContrib) %>% 
  select(pathways,
         MGH03D.related,
         Other,
         rel_abund.highest,
         importance)

output_name = 'sPLSDA.MGH03D.Functional.Loadings'
write_tsv(pL, paste0(results_dir, fig, output_name, '.tsv'))
```

```{r}
output_name = 'sPLSDA.MGH03D.Functional.Heatmap'
mixOmics::cim(data.splsda, 
              symkey = T, keysize = c(0.8, 0.8), 
              row.names = F,
              row.sideColors = sample_colors$color, 

              col.names = gsub(data.splsda$names$colnames$X, pattern = '_noname', replacement = ''),
              save = 'pdf', 
              name.save = paste0(results_dir, fig, output_name),
              margins = c(33, 15), row.cex = .35, col.cex = .35,
              # margins = c(25,20), row.cex = .8, col.cex = .8,
              
              transpose = T,
              # row.cex = .5, col.cex = .75,
              # margins = c(6.5, 20), 
              scale = F, center = T)
```

